---
title: "Introduction to hgnc"
author: "Zhuoer Dong"
date: "`r Sys.Date()`"
output: 
    prettydoc::html_pretty:
        toc: true
        number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to hgnc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = F, comment = '', collapse = TRUE)

library(magrittr)
```

```{r}
library(hgnc)
```



# quick start

For impatient people, you can convert a probe-ensembl map like the following [^probe2ensembl]

```{r}
probe2ensembl <- tibble::tribble(
  ~ID,           ~Ensembl,                                                 
  "200064_at",   "ENSG00000096384",                                        
  "200066_at",   "ENSG00000113141",                                        
  "200068_s_at", "ENSG00000127022",                                        
  "200069_at",   "ENSG00000075856",                                        
  "200071_at",   "ENSG00000119953",                                        
  "200076_s_at", "ENSG00000105700",                                        
  "200077_s_at", "ENSG00000104904",                                        
  "200078_s_at", "ENSG00000117410",                                        
  "200082_s_at", "ENSG00000171863 /// ENSG00000183405 /// ENSG00000213326",
  "200084_at",   "ENSG00000110696"
)

probe2ensembl
```

[^probe2ensembl]: how I create `probe2ensembl`
    
    ```r
    # read probe annotation from a GSE SOFT file
    soft_table <- system.file("extdata/GSE19161_family.soft.gz", package = "rGEO") %>% 
        rGEO::parse_gse_soft(verbose = F) %>% {.$table}
    # subset part of the table for the purpose of demostration
    probe2ensembl <- soft_table %>% dplyr::select(1, 9) %>% dplyr::slice(41:50)
    ```

to a probe-symbol map ^[Ensembl IDs which don't have corresponding HUGO symbol are discarded.]

```{r}
probe2_symbol <- probe2ensembl %>% melt_map("ID", "Ensembl", " /// ") %>%
    dplyr::mutate("symbol" = as_symbol_from_ensembl(Ensembl)) %>%
    cast_map("ID", "symbol", " /// ")

probe2_symbol
```

Other common ID like entrez gene ID, Unigene ID, RefSeq accession are also supported.



# innovation

## the problem

```{r}
probe2id <- tibble::tribble(  
	~probe, ~id,
    "probe1", "id1 | id2",
    "probe2", "id3"
)

probe2symbol <- tibble::tribble(  
	~probe, ~symbol,
    "probe1", "symbol1 | symbol2",
    "probe2", "symbol3"
)
```

To master this package, you need to understand the problem it aims to solve. That is, to turn something like

```{r}
probe2id
```

to 

```{r}
probe2symbol
```



## solve a simpler one

```{r}
probe2id_easy <- tibble::tribble(  
	~probe, ~id,
    "probe4", "id4",
    "probe5", "id5",
    "probe6", "id6"
)

probe2symbol_easy <- tibble::tribble(  
	~probe, ~symbol,
    "probe4", "symbol4",
    "probe5", "symbol5",
    "probe6", "symbol6"
)
```

To point out the key difficulty, let's contrast it with a simpler one --- to turn something like 

```{r}
probe2id_easy
```

to 

```{r}
probe2symbol_easy
```

That's quite easy, you just need a id-symbol map, 

```{r}
id2symbol <- tibble::tibble(  
    id = paste0("id", 1:6),
    symbol = paste0("symbol", 1:6)
) %>% dplyr::sample_frac()

id2symbol
```

and use the following code ^[I deliberately shuffle the rows of `id2symbol` to show that `id2symbol` just need to provide the correct relationship between `id` and `symbol`, i.e, it doesn't necessarily maintain the same order as `probe2id`.]

```{r}
dplyr::transmute(probe2id_easy, probe, symbol = id2symbol$symbol[match(id, id2symbol$id)])
```

In the above code, we map `probe` to `symbol` in three steps:

1. `dplyr::transmute` preserves `probe2id$probe` - `probe2id$id` relationship by _position_
1. `match()` finds `probe2id$id` - `id2symbol$id` relationship by _value_
1. `[]` finds `id2symbol$id` - `id2symbol$symbol` relationship by _position_

Let's us understand the example by a concrete example of `"probe4"` to `"symbol4"`:

1. 1st element of `probe2id$probe` -> 1st element of `probe2id$id`:  
   `"probe4"` is the 1st element of `probe2id$probe`, so we look for the 1st element of `probe2id$id`, `"id4"`.

1. `"id4"` in `probe2id$id` -> `"id4"` in `id2symbol$id`:  
    `"id4"` is the 1st element of `probe2id$id`, then we look for the 1st element of `match()` (which gives the position of `probe2id$id` in `id2symbol$id` --- ``c(`r match(probe2id$id, id2symbol$id)`)`` in this case). We get `3`, so we look for the 3rd element of `id2symbol$id`, the exact value of `"id4"`.

1. 3rd element of `id2symbol$id` -> 3rd element of `id2symbol$symbol`:  
    finally, `"id4"` is the 3rd element of ``id2symbol$id`, thus we look for the 3rd element of `id2symbol$symbol`, `"symbol4"`.



## key difficulty 

Back the original problem, you can find that its fairly easy to "replace" `"id4"` with `"symbol4"`, `"id5"` with `"symbol5"`, etc (thanks to `match()`). But how can you "replace" the `"id1"` and `"id2"` inside `"id1 | id2"`? 

That is what we meet exactly, as in the 9th line of `probe2ensembl`.

```{r}
probe2ensembl %>% dplyr::slice(9)
```

If you think it's a piece of cake, you may have some misunderstanding: 

- Computer is very foolish, it can't convert `"id1 | id2"` to `"symbol1 | symbol2"` as you can easily achieve even without thinking. In programming, the only way is to search `"id1"` in `"id1 | id2"` and replace with `"symbol"` if it find one, then search `"id2"`, `"id3"`, etc. This will cause a severe poor performance.

- As for replacing all `"id"` with `"symbol"`, I use `id2symbol` just to simplify the problem, the id-symbol map in the real world is usually something like:

  ```{r}
  hgnc::ensembl2symbol %>% dplyr::sample_n(3)
  ```

## solution

Inspired by [**reshape2**](https://github.com/hadley/reshape), I choose to **melt** the **wide** map

```{r}
probe2id_wide <- probe2id

probe2id_wide
```

to a **long** map.

```{r}
probe2id_long <- probe2id_wide %>% melt_map("probe", "id", " \\| ")

probe2id_long
```


Then map id to symbol to get a new **long** map, following the way we solve the simpler problem abobe.

```{r}
probe2symbol_long <- probe2id_long %>% 
    dplyr::transmute(probe, symbol = id2symbol$symbol[match(id, id2symbol$id)])

probe2symbol_long
```

Finally **cast** to a new **wide** map

```{r}
probe2symbol_wide <- probe2symbol_long %>% cast_map("probe", "symbol", " /// ")

probe2symbol_wide   # now it's identical to probe2symbol
```

In short, A-B wide map -> A-B long map -> A-C long map -> A-C wide map.



# real world

In above discussion, I abstract away many details to focus on core idea. Things get more complicated in real world:

- The separator is not definite (it can be ` `, ` 、`, ` \\\ `), so I use [reverse match](https://github.com/dongzhuoer/thesis/r-lib/rGEO/blob/master/R/platform.R#L184-L189).
- I omit where `id2symbol` comes from (at least not falls from sky). Actually I **melt** [hgnc_complete_set.txt.gz](https://github.com/dongzhuoer/thesis/r-lib/hgnc/blob/master/data-raw/hgnc_complete_set.txt.gz) to create `entrez2symbol`, `ensembl2symbol`, etc.
- I use `symbol = id2symbol$symbol[match(id, id2symbol$id)]` in the above code for universality, but it looks quiet obscured even though I have explained in great detail. Thus I add some syntax sugar (`?as_symbol`), so that you can see the simple and beautiful code in the beginning.

Armed by with above **weapon**, the package can serves as the workhorse of [**rGEO**](https://github.com/dongzhuoer/thesis/r-lib/rGEO) to transform any user-supplied GPL file to standard chip file ready for GSEA.
