---
title: processing raw data to package data
date: "`r Sys.Date()`"
---

```{r}
knitr::opts_knit$set(root.dir = here::here(''))
```



# clean

```{r clean, include=FALSE}
dir(here::here('data'), full.names = T) %>% file.remove()
rm(list = ls(envir = globalenv(), all = T))
```



# Process raw data

If you find find something puzzling, refer to `R/platform.R`.

```{r setup, results='hide', echo=FALSE}
knitr::opts_chunk$set(error = T, collapse = TRUE)

options(tibble.print_max = 40, tibble.width = Inf)
```

```{r}
library(tidyverse)
library(parallel)

pkgload::load_all()
```

## prepare GPL metadata

```{r}
gpl_metas <- rGEO.data::gpl_metas
n_all <- length(gpl_metas)

sum_length <- . %>% sapply(length) %>% sum
# print selected gpl_metas
detail <- function(x, info_only = F) {
	if (info_only)
		print(gpl_metas[unique(x$accession)] %>% lapply(. %>% {.$info}))
	else
		print(gpl_metas[unique(x$accession)])
}
```

## remove null

```{r}
gpl_null <- gpl_metas %>% {.[sapply(., . %>% {.$info} %>% is.null)]} %>% names %T>% {print(head(.))}

gpl_metas %<>% {.[setdiff(names(.), gpl_null)]}
```

## deal with duplicated columnname

```{r cache=T}
# for all cases in which there is duplicated name, remove those name and description are both same, then manually check left case
gpl_metas %>% lapply(. %>% {.$info}) %>% 
	lapply(. %>% {
		if (.$name %>% duplicated %>% any) {
			. <- .[paste(.$name, .$description) %>% {!duplicated(.)}, ];
		}
		.
	}) %>% {.[sapply(., . %>% {duplicated(.$name)} %>% any)]} %>% 
	lapply(. %>% filter(name %in% name[duplicated(name)])) %>%
	yaml::write_yaml('tests/testthat/output/gpl-dup-col-name.yaml', column.major = F)

message('search "AlleleA_ProbeSeq in description" or "Illumina Strand 2" in description if you want to use sequence information')

#" for these special accession, the second one is better (refseq or entrez)
gpl_metas[dup_use_latter] %<>% 
	{print(lapply(., . %>% {.$info} %>% filter(name %in% name[duplicated(name)]))); .} %>%
	lapply(function(gpl) {
		duplicated_name <- gpl$info$name %>% {.[duplicated(.)]};
		col_drop <- sapply(duplicated_name, . %>% {which(gpl$info$name == .)[1]})
		gpl$info   = gpl$info[-col_drop, ];
		gpl$sample = gpl$sample[-col_drop];
		
		gpl
	})

#" for others, simply pick first is enough
gpl_metas %<>% lapply(function(gpl) {
	dup <- duplicated(gpl[[1]]$name)
	if (dup %>% any) {
		gpl$info   = gpl$info[!dup, ];
		gpl$sample = gpl$sample[!dup]
	}
	
	gpl
}) 

assertthat::assert_that(
	gpl_metas %>% lapply(. %>% {colnames(.$sample) %>% {.[duplicated(.)]}}) %>%
		unlist %>% unique %>% libzhuoer::print_or_T(),
	msg = 'duplicated column names not dealt'
)
```

## make info df to be compatible with dplyr API

```{r make-info}
info_all <- setdiff(names(gpl_metas), special$accession) %>%
	mclapply(. %>% {add_column(gpl_metas[[.]]$info, accession = ., .before = 1)}) %>%
	bind_rows() %>% mutate(full = paste0(name, ': ', description)) %>%
	add_column(type = 'unknown') %T>% print
```

## different types of GPL

### non human
```{r}
gpl_non_human <- info_all %>% filter(type == 'unknown') %>% filter_non_human() %>% 
	filter(!duplicated(accession)) %>% {.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_non_human, 'non_human', type))
```

###  ncRNA
```{r}
gpl_miRNA <- info_all %>% filter(type == 'unknown') %>% filter_miRNA() %>% 
	filter(!duplicated(accession)) %>% {.$accession} %T>% {print(head(.))} %T>% {print(head(.))}
info_all %<>% mutate(type = ifelse(accession %in% gpl_miRNA, 'miRNA', type))

gpl_circRNA <- info_all %>% filter(type == 'unknown') %>% filter_circRNA() %>% 
	filter(!duplicated(accession)) %>% {.$accession} %T>% {print(head(.))} %T>% {print(head(.))}
info_all %<>% mutate(type = ifelse(accession %in% gpl_circRNA, 'circRNA', type))
```

### entrez id
```{r}
gpl_entrez_id <- info_all %>% filter(type == 'unknown') %>% filter_entrez_id() %>%
	{.$full %>% unique %>% cat(sep = '\n'); .} %>%
	{.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_entrez_id, 'entrez_id', type))
```

### entrez
```{r}
gpl_entrez <- info_all %>% filter(type == 'unknown') %>% filter_entrez %>% 
    {.$full %>% unique %>% cat(sep = '\n'); .} %>%
	{.$accession} %T>% {print(head(.))}

message('we don\'t distinguish between id and symbol here')

info_all %<>% mutate(type = ifelse(accession %in% gpl_entrez, 'entrez', type))
```

### ORF
```{r}
gpl_ORF <- info_all %>% filter(type == 'unknown') %>% filter_ORF() %>%  
	{.$full %>% unique %>% cat(sep = '\n'); .} %>%
	{.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_ORF, 'ORF', type))
```

### symbol
```{r} 
gpl_symbol <- info_all %>% filter(type == 'unknown') %>% filter_symbol() %>%
	{.$full %>% unique %>% cat(sep = '\n'); .} %>%
	{.$accession} %T>% {print(head(.))}

message('For those whose sample is all empty, we still can\'t exclude the possibility that the field is name or description rather than symbol')

info_all %<>% mutate(type = ifelse(accession %in% gpl_symbol, 'symbol', type))
```

### ensembl
```{r}
gpl_ensembl <- info_all %>% filter(type == 'unknown') %>% filter_ensembl() %>%
	{.$full %>% unique %>% cat(sep = '\n'); .} %>% 
	{.$accession} %T>% {print(head(.))} 

info_all %<>% mutate(type = ifelse(accession %in% gpl_ensembl, 'ensembl', type))
```

### refseq
```{r}
gpl_refseq <- info_all %>% filter(type == 'unknown') %>% filter_refseq() %>% 
	{.$full %>% unique %>% cat(sep = '\n'); .} %>%
	{.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_refseq, 'refseq', type))
```

### genbank
```{r}
gpl_genbank <- info_all %>% filter(type == 'unknown') %>% filter_genbank() %>%  
	{.$full %>% unique %>% cat(sep = '\n'); .} %>% 
	{.$accession} %T>% {print(head(.))}
#" at least for GPL10348, name is better

info_all %<>% mutate(type = ifelse(accession %in% gpl_genbank, 'genbank', type))
```

### unigene
```{r}
gpl_unigene <- info_all %>% filter(type == 'unknown') %>% filter_unigene() %>% 
	{.$full %>% unique %>% cat(sep = '\n'); .} %>% 
	{.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_unigene, 'unigene', type))
```

### sequence
```{r}
gpl_sequence <- info_all %>% filter(type == 'unknown') %>% filter(type == 'unknown') %>% filter_sequence() %>% 
	{.$full %>% unique %>% cat(sep = '\n'); .} %>% 
	{.$accession} %T>% {print(head(.))}

info_all %<>% mutate(type = ifelse(accession %in% gpl_sequence, 'sequence', type))
```

### code stored for future ID type
```{r}
	# {.$full %>% unique %>% print; .} %>%
	# filter(!str_detect(full, fixed('description', T))) %>%  
	# {filter(., str_detect(description, fixed('alias', T))) %>% detail; .} 

	# lapply(. %>% {tibble(accession = ., symbol = select(gpl_metas[[.]]$sample, contains('symbol'))[[1]])}) %>% bind_rows() %>% filter(symbol != '') %>% filter(!(symbol %in% hgnc::hugo_symbol_all)) %>% group_by(accession) %>% summarise(n = n()) %>% arrange(desc(n))
```

### unknown
```{r}
gpl_unknown <- info_all %>% filter(type == 'unknown') %>% {unique(.$accession)} %T>% {print(head(.))}
```

## check for possible omission

```{r}
assertthat::assert_that(
	n_all == list(gpl_null, special, unique(info_all$accession)) %>% sum_length,
	msg = 'Not all platforms analysed'
)

assertthat::assert_that(
	full_join(
		ls(pattern = '^gpl_') %>% {.[. != 'gpl_null']} %>% 
			{tibble(type = ., n = sapply(., . %>% get %>% length))} %>% 
			mutate(type = str_remove(type, '^gpl_')), 
		info_all %>% filter(!duplicated(accession)) %>% count(type) %>% ungroup(),
		by = 'type'
	) %>% filter(n.x != n.y) %>% libzhuoer::print_or_T(),
	msg = 'some IDs are parsed incorrectly'
)
```

## Sum up & use data

```{r}
gpl <- bind_rows(
	info_all %>% filter(!duplicated(accession)) %>% select(accession, type), 
	select(special, accession, type = database), 
	tibble(accession = gpl_null, type = 'unknown')
)

assertthat::assert_that(
	n_all == nrow(gpl),
	msg = 'not all platforms included in `gpl` data'
)

usethis::use_data(gpl, overwrite = T)
```


# Afterward

```{r reinstall}
devtools::test()     # test the new data
roxygen2::roxygenize() # you may also have edited data documentation

system('R CMD INSTALL --no-multiarch --with-keep.source .')
devtools::reload()   # now you can use the new data in current R session 
```







