---
title: "Introduction to rGEO"
author: "Zhuoer Dong"
date: "`r Sys.Date()`"
output: prettydoc::html_pretty
vignette: >
  %\VignetteIndexEntry{Introduction to rGEO}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

options(tibble.print_min = 6, tibble.max_extra_cols = 1)
```

```{r}
library(rGEO)
```



# quick start

You can find raw data at https://www.ncbi.nlm.nih.gov/geo/browse/ ^[search for 
    `GSE***`, download files from the `Download family` section at the bottom.],
    or use the following handy tools:
    
```{r}
gse_soft_ftp('GSE51280')

gse_matrix_ftp('GSE51280')
```
    
Then you can read them:

```{r}
# expression matrix file
read_gse_matrix(system.file('extdata/GSE51280_series_matrix.txt.gz', package = 'rGEO'))

# SOFT (chip annotation) file
read_gse_soft(system.file('extdata/GSE51280_family.soft.gz', package = 'rGEO'))
```



# A little deeper

Let's go a little deeper to see how this works. 

The package mainly contains functions for   
   1. telling the ftp path to GSE raw data: see `gds_ftp()` and `gpl_soft_ftp()`  
   1. reading these files, including expression matrix file and SOFT (chip 
        annotation) file, see below.  

## expression matrix file

reading expression matrix file is quiet straightforward, you just need to locate
    where the information lies, and `readr::read_tsv()` would take care of the 
    rest for you.

```{r eval=FALSE}
read_gse_matrix
```

```{r results='asis', echo=FALSE}
cat('```r\n')
read_gse_matrix
cat('```\n')
```



## SOFT (chip annotation) file

SOFT (chip annotation) file is a lot more harder to deal with. In short, you 
    have to dig useful information first:
    
```{r}
parse_gse_soft(system.file('extdata/GSE51280_family.soft.gz', package = 'rGEO'), F)
```

and convert that to a uniform format. Almost the whole package does the latter, 
    refer to `vignette('probe2symbol')` for more details. 

By the way, for both `parse_gse_soft()` and `read_gse_soft()`, you can add 
    `verbose = T` to make the process more clear.



# Innovation

> The format of SOFT (chip annotation) file is really a terrible messy, or rather, aganist humanity. Even though I have developed this handy package, I never want to touch it again.

The central problem is that we don't know each probe corresponding to which gene. 

The annotation (TSV) contains various information, such as Entrez Gene ID、Ensembl gene ID、INSDC accession、RefSeq accession (nucleotide), so I developed [hgnc](https://github.com/dongzhuoer/thesis/r-lib/hgnc) to universely convert them to [HUGO](https://www.genenames.org/) gene symbol. That is actually a fairly easy task, what real drives one crazy is the messy column names. Given the following snippet ^[the left part is column name, and the right part is explanation], how can you make sure which columns exactly store Entrez Gene ID? ......

```
EntrezGeneID: EntrezGene ID
ORF: Entrez Gene identifier for the gene represented
GeneID: Entrez Gene identification number
UniGene_ID: UniGene Database Accession Number (http://www.ncbi.nlm.nih.gov/sites/entrez?db=unigene)
ORF: Entrez GENE ID
Entrez_Gene_ID: Entrez gene ID
GENE: Entrez GeneID of sequence used to design cDNA probe
Gene_ID: id in NCBI Entrez Gene database
```
